A family of accessibility measures and the concept of potential derived from spatial interaction principles

### 2025-04-30 paper/manuscript.qmd
Another revision of the draft done --- off to AP!

---
title: "A family of accessibility measures derived from spatial interaction principles"
format:
  pdf:
    header-includes:
        - \usepackage{float}
        - \floatplacement{table}{H}
        - \floatplacement{figure}{H}
bibliography: "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
#includes:
#      in_header: 'gt_packages.sty'
      #\usepackage{float}
---

@@

<!-- SUMMARY: Accessibility is related to mobility, however, institutionalized mobility-based transportation planning practice has treated accessibility as it's proxy. This has had problematic consequences. Conceptually, accessibility is a more complete and comprehensive evolution of mobility. As such, there has been recognition that the adoption of accessibility planning approaches are needed; however there are 1) too many definitions and 2) uncertainty in meaning of the outputs-->

Historically, the focus of modern transportation planning has been to prioritize mobility while treating access to destinations as a by-product of movement. This has had problematic consequences: with the car seen as the ultimate mobility tool, this approach has led to the emergence and continued dominance of an automobility mono-culture [@miller_collaborative_2011; @lavery_driving_2013]. Decades of planning for this mono-culture (often characterized by road and highway expansion) have been marked by increased travel cost and environmental burdens, but often having only limited impact on the ease with which people can reach destinations [@farber_running_2011; @handyACCESSIBILITYVSMOBILITYENHANCING2002; @paez_healthcare_2010]. In response to this situation, transportation researchers have increasingly advocated for the adoption of accessibility as a planning criterion [@silvaAccessibilityInstrumentsPlanning2017; @paez_developing_2013; @handy2020] using both positive and normative approaches [@paez2012measuring; @levineCenturyEvolutionAccessibility2020]. Accessibility, a central concept in transport geography, planning, and engineering, is conceptually related to human mobility and the opportunity landscape. In simple terms, it is defined as the "potential of opportunities for interaction" [@hansen1959]. Compared to other measures of performance used in transportation that benchmark movement (e.g., VKT, PKT, etc.), accessibility brings a more holistic understanding of transportation and land use systems combined [@handyMeasuringAccessibilityExploration1997].

<!-- I suggest removing 'using both positive and normative approaches [@paez2012measuring; @levineCenturyEvolutionAccessibility2020]' from the paragraph above -->

<!-- the first two paragraphs set the context / expectaction for the paper. In the current form, it the first two paragraphs seem to be more focused on the importance of accessibility for planning, and the challenges about adopting accessibility into planning practice. However, the paper is very much a theoretical and methodological contribution. the first paragraphs should reflect this, drawing attention to the common origin of spatial interaction models and accessibility , and how they diverged over time and we got lost along the way . Some pepople might not have a clear understanding of how the two concepts connect or how exactly they differ. this could be briefly explained -->

A growing interest in accessibility has been accompanied by a remarkable boom in scholarly research, which has grown to include studies of access to employment [e.g., @karstEvaluationAccessibilityImpacts2003; @grengs2010job; @paez_jobs_2013; @merlin2017competition; @tao_investigating_2020], health care [e.g., @luo2003; @paez_healthcare_2010; @wan2012three; @delamater2013spatial; @boisjoly2017informality; @pereira_2021_geographic], green spaces [@reyesAccessibility2014; @rojas_accessibility_2016; @liang_novel_2024], schools [e.g., @williams_disparities_2014; @romanillosAccessibilitySchoolsSpatial2018; @marques_accessibility_2021], social contacts [e.g., @neutens_human_2007; @farberActivitySpacesMeasurement2012; @farber_2013_social], and regional economic analysis [e.g., @vickermanAccessibility1999; @lopezMeasuring2008; @ribeiro_road_2010; @gutierrez_evaluating_2011] among many other domains of application. In other words, accessibility analysis is used today to broadly understand the potential to reach opportunities that are important to people [@ferreiraReenactingMobilityAccessibility2020]. However, despite its growth in popularity in scholarly works, challenges remain with respect to the more widespread adoption of accessibility in planning practice. Several barriers to bridging the scholarly-practice accessibility gap have been identified. For example, the diversity of accessibility definitions has been flagged by @vanweeAccessible2016, @handy2020, and @kapatsila_resolving_2023. Further, difficulties in the interpretability and communicability of outputs has also been noticed by @geursAccessibilityEvaluationLanduse2004, @vanweeAccessible2016, and @ferreiraReenactingMobilityAccessibility2020.

The adoption of accessibility in planning practice is not necessarily made easier when potential adopters have to contend with a plethora of definitions, each seemingly more sophisticated but less intuitive than the last [@kapatsila_resolving_2023]. At a high level, @geursAccessibilityEvaluationLanduse2004 identify four families of accessibility measures: infrastructure-, place-, person-, and utility-based. Of the place-based family, which is the focus of this paper, the menu has grown to include gravity-based accessibility [e.g., @hansen1959; @pirie_measuring_1979], cumulative opportunities [e.g., @wachs_physical_1973; @pirie_measuring_1979; @ye_spatial_2018], modified gravity [e.g., @SchuurmanMeasuring2010], 2-Step Floating Catchment Areas [e.g., @luo2003], Enhanced 2-Step Floating Catchment Areas [e.g., @luoEnhanced2009], 3-Stage Floating Catchment Areas [e.g., @wan2012three], Modified 2-Step Floating Catchment Areas [e.g., @delamater2013spatial], inverse 2-Step Floating Catchment Areas [e.g., @wang_2sfca_2021], and n-steps Floating Catchment Areas [@liang_novel_2024]. How is a practitioner to choose among this myriad options? What differences in accessibility scores should matter, and how should they be communicated? [see @vanweeAccessible2016; p. 14].

In this respect, @wuUnifyingAccess2020's contribution provides a unifying framework to think about accessibility measures, the most general of which they describe as $S_i$ in @eq-access-Lu-Levinson. By considering the key features of accessibility, namely travel-cost $f(c_{ij})$ and the distribution of opportunities $g(O_i)$, @wuUnifyingAccess2020 demonstrate that a majority of the concepts found scattered throughout the accessibility literature can be seen as particular cases of a general accessibility formula (@eq-access-Lu-Levinson). One needs only to judiciously change the way travel costs and opportunities are formulated to derive almost any known place-based accessibility measure. In this way, @wuUnifyingAccess2020 have taken a considerable step towards clarifying the differences between various accessibility indicators.

<!--  the paragraph above is really good -->

$$
S_i \propto \sum_j g(O_j)f(c_{ij})
$$ {#eq-access-Lu-Levinson}

However, there is room to further clarify another relevant aspect: competition, i.e. when finite resources can be concurrently accessed by many people. @wuUnifyingAccess2020's unifying framework does not appear to accommodate many (or any) of the floating catchment area methods, a popular approach in the accessibility literature that accounts for competition. There have also been more recent developments in the literature: for instance, @soukhovIntroducingSpatialAvailability2023 demonstrate that the introduction of a single constraint is sufficient to turn the general accessibility measure (i.e, @eq-access-Lu-Levinson) into a competitive measure of accessibility in accordance with the 2-Step Floating Catchment Area approach. Moreover, the proportional allocation balancing factors in @soukhovIntroducingSpatialAvailability2023 apply a singly-constraint to the accessibility measure, akin to the constraints introduced within the spatial interaction modelling framework outlined in @wilson1971.

<!-- Most readers will not be familiar with your previous papers, so they will not understand what you mean with 'single constraint' or 'balancing factors'. These ideas need to be explained in this paper. I suggest the text 'There have also been more recent.....' in the paragraph is moved to the literature review section, where we have proper space to explain those ideas and how the ideas proposed in this paper expand on then -->

Motivated by the challenge of incorporating competition into a unified framework for accessibility, this paper contends that accessibility research must reconnect with its spatial interaction origins. Particularly, we argue that an important aspect of spatial interaction modelling--namely, constraining the results to match empirical observations--was never effectively reincorporated into accessibility analysis. Empirical constraints were embraced by early spatial interaction literature following the work of @wilson1971, but this stream of literature tended to flow separately from research inspired by @hansen1959' accessibility. The application of @wilson1971's empirical constraints supported the development of various spatial interaction models that remain relevant in research and practice today [@ortuzar_2011_modelling]. However, the same cannot be said of the contemporary accessibility literature, where empirical constraints were not explicitly adopted. We argue that the absence of empirical constraints (and their attendant proportionality constants) has contributed to some of accessibility analysis' interpretability issues; for instance, the fuzziness of insights beyond simple proportional statements like 'higher-than' or 'lower-than' [@millerAccessibilityMeasurementApplication2018].

<!-- when you say that the absence of empirical constraints has led to interpretation issues in accessibility analysis, and think the example should be stronger that simply saying it leads to the 'fuzziness of insights' . perhaps mention the problem that we loose track of measurement units, that comparability between indicators, across cities and transport modes also become compromised but most of us don't even notice it, and this should be explained in the next sections -->

These streams of literature share common headwaters. It is by looking to the past that we believe accessibility analysis can newly wade into the future. Hence, this work's primary focus is to show that the same empirical constraints used in @wilson1971's family of spatial interaction models can be mapped onto accessibility (REITERATE WHY). We begin by tracing the development of accessibility from its origins in spatial interaction, from the Newtonian gravitational expression in @ravensteinLawsMigration1889 through to the seminal accessibility work of @hansen1959. We then present evidence for a narrative highlighting the marked divergence between accessibility and spatial interaction modelling research after the work of @wilson1971. Next, we hark back to @wilson1971's spatial interaction models, and use it to derive a family of accessibility measures based on different types of constraints. We illustrate various members of this family with a simple numerical example and a real world data set. We then conclude by discussing the uses of these measure and their interpretation.

```{=html}
<!--
We wish to demonstrate the importance of transiting from proportionality to equality in @eq-accessibility-general, in order to balance the units of accessibility into more interpretable values, as well as converting accessibility indicators from ordinal variables into interval/ration ones. To address this matter, we need not reinvent the wheel. Stewart left the determination of $G$ for the future, possibly expecting it to be some sort of universal constant. However, the solution to this problem was found by @wilson1971, who discovered that there was not a single $G$, but rather various balancing factors that depend on 1) empirical observations of population and/or opportunities; and 2) whether (and how) they were used to constrain the analysis. This development is outlined next.
-->
```

@@

To be certain, the expression above, a proportionality, is one of the most famous in all of science. In brief, it states that the force of attraction $F$ between a pair of bodies $i$ and $j$ is directly *proportional* to the product of their masses $M_i$ and $M_j$, and inversely *proportional* to the square of the distance between them $D_{ij}$. Direct proportionality means that as the product of the masses increases, so does the force. Likewise, inverse proportionality means that as the distance increases, the force decreases. @eq-phys-grav-prop, however, does not quantify the magnitude of the force. To do so, an empirical constant, aka gravitational constant, is required to convert the proportionality into an equality, ensuring that values of the force $F$ in @eq-phys-grav-prop match the observed force of attraction between masses. In other words, @eq-phys-grav-prop needs to be *constrained* using empirical data. Ultimately, the equation for the force is as seen in @eq-phys-grav, where $G$ is an empirically calibrated proportionality constant:

@@

# Early research on human spatial interaction: from Ravenstein (1889) to Stewart (1948)

```{=html}
<!-- 
SUMMARY: a review of spatial interaction literature from the 1880s to the 1940s. A number of researchers theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $Mi$ and $Mj$ and inversely proportional by their separation distance. This concept was captured with different expressions, but all tie back to the same Newtonian gravity analogy, although not all of them included a proportionality constant in their formulation.
```

Since the 1880s to the 1940, a number of researchers theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $Mi$ and $Mj$ and inversely proportional by their separation distance. This concept was captured with different expressions, but all tie back to the same Newtonian gravity analogy, although not all of them included a proportionality constant in their formulation.

@@

-   $F$ is the *demographic force*
-   $N_1$ and $N_2$ are the numbers of people of in groups 1 and 2
-   $\mu_1$ and $\mu_2$ are so-called *molecular weights*
-   $M_1 = \mu_1N_1$ and $M_2 = \mu_2N_2$ are the demographic masses at 1 and 2
-   $d_{12}^2$ is the distance between $1$ and $2$
-   And finally $G$, a constant that Stewart "left for future determination" [-@stewartDemographicGravitationEvidence1948, p. 34]

<!-- what exactly did he mean by 'molecular weights' ? -->

In addition to demographic force, Stewart defined a measure of the "population potential" of $2$ with respect to $1$ as follows:

<!-- in plain language, what do you mean by 'the population potential of 2 with respect to 1' ? the potential number of people from location 2 that would visit location 1 ? but also, I'm not sure we need this equation 7 -->

\
$$
V_1 = G\frac{M_2}{d_{12}}
$$ {#eq-stewart-population-potential}

For a system with more than two population bodies, Stewart formulated the population potential at $i$ as follows (after arbitrarily assuming that $G=1$): $$
V_i = \int\frac{D}{r} ds
$$ {#eq-stewart-population-potential-integral}

\noindent where $D$ is the population density over an infinitesimal area $ds$ and $r$ is the distance to $i$. In @eq-stewart-population-potential-integral, $D\cdot ds$ gives an infinitesimal count of the population, say $dm$, and so, after discretizing space, @eq-stewart-population-potential-integral can be rewritten as: $$
V_i = \sum_j \frac{M_j}{d_{ij}}
$$ {#eq-stewart-population-potential-sum}

@@

<!--I say we spell it out as in this eq^, what you think?-->

@@

# Hansen's gravity-based accessibility to today

@@

However, Hansen' use of Stewart's *population potential* measure included one crucial omission that afflict the literature to this day. The omission is that that between @stewartDemographicGravitationEvidence1948 and @hansen1959 the proportionality constant $G$ in @eq-stewart-population-potential vanished. This constant was not explicitly addressed in @hansen1959 and accessibility research continues to evolve without it. Since @hansen1959, accessibility analysis has been widely used in numerous disciplines but, to our knowledge, the proportionality constant has remained implicit or forgotten, with no notable developments to explicitly determine it.

Paper, the omission of this constant generates a fundamental problem for the measurement unit of accessibility estimates, which has fundamental implications for the interpretation, communication and comparability of accessibility analysis. At any rate, those reading @hansen1959 must recall that @stewartDemographicGravitationEvidence1948 had set the proportionality constant $G$ to 1, with a note that "$G$ \[was\] left for future determination: a suitable choice of other units can reduce it to unity". In practice, the persistent omission of the costant in accessibility analyses  means that $G$ continues to be implicitly set to $1$, even when the fundamental relationship in accessibility is proportionality (e.g., $S_{i}  \propto \sum_j g(O_j)f(d_{ij})$) and not equality [for instance, see the formula for accessibility at the top of Figure 1 in @wuUnifyingAccess2020]. The direct consequence is that without a proportionality constant, the units of $S_i$ remain unclear: the unit of "potential of opportunity for interaction" is left free to change as $\beta$ is calibrated. For example, if $c_{ij}$ is distance in meters, it will be number of opportunities per $m^{\beta}$ when $f(c_{ij}) = d^{-\beta}$ but number of opportunities per $e^{-\beta\cdot m}$ when $f(c_{ij}) = e^{\beta\cdot m}$. This undermines the comparability of accessibility metrics with different decay functions, and renders their results difficult to understand and communicate. Hansen-style accessibility, therefore, is better thought of as an ordinal measure of potential that can only be interpreted in terms of higher and lower accessibility, but which has not palpable meaning [@millerAccessibilityMeasurementApplication2018].

<!-- jargon ? 'a suitable choice of other units can reduce it to unity' I think this could be explained in plain English -->

@@

<!-- very brief narrative to introduce - is this happening separately from hansen? "On the other side of the atlantic, wilson was doing..." :) -->

<!-- rafa: I agree it would be nice to have a sentence or two that sets the context of wheter wilson was aware of stewart's and hansen's work -->

In his groundbreaking work, Wilson [-@wilson1971] defined a general spatial interaction model as follows:

@@

The model in @eq-phys-gravity-model posits a quantity $T_{ij}$ that represents a value in a matrix of flows of size $n \times m$, that is, between $i = 1,\cdots, n$ origins and $j = 1,\cdots, m$ destinations. The quantities $W_i^{(1)}$ and $W_j^{(2)}$ are proxies for the masses at $i=1,\cdots,n$ origins and $j=1,\cdots,m$ destinations. Finally, $f(c_{ij})$ is some function of travel cost $c_{ij}$ which reflects travel impedance. In this way, $T_{ij}$ explicitly measures *interaction* in the unit of trips, and the role of $k$ is to ensure that the system-wide sum of $T_{ij}$ represents the total flows in the data. In other words, $k$ is a scale parameter that makes the overall amount of flows identical to the magnitude of the phenomenon being modeled. In other words, it balances the units, just like the gravitational constant in Newton’s Law of Universal Gravitation.

@@

Additional information can be introduced. For example, when information is available about the total number of trips produced by each origin ($O_i$), the following constraint can be used:

>>>>

---
title: "A family of accessibility measures and the concept of potential derived from spatial interaction principles"
format:
  pdf:
    header-includes:
        - \usepackage{float}
        - \floatplacement{table}{H}
        - \floatplacement{figure}{H}
bibliography: "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
#includes:
#      in_header: 'gt_packages.sty'
      #\usepackage{float}
---

@@

<!-- SUMMARY *AND OUTLINE FOR ABSTRACT*: Transportation planning, historically, has been mobility-oriented. Accessibility, however, should be the primary focus, and institutionalized mobility-based transportation practice has treated accessibility as its proxy. This has had problematic consequences. Conceptually, accessibility is a more complete and comprehensive evolution of mobility: from movement, to the potential movement to destinations. As such, there has been recognition that the adoption of accessibility planning approaches are needed; however there are  too many definitions and uncertainty in meaning of the outputs. This work's goal is to enhance interpretation of values, clarify the concept of accessibility, and unify other measures of accessibility such as the popular Hansen-type measure the 2SFCA.

To do so, this work demonstrates that accessibility and mobility oriented models (i.e., gravity-based accessibility and the spatial interaction model) share the same roots in spatial interaction modelling, and a family of accessibility measures can be specified within the spatial interaction modelling principles. In demonstrating these developments, three significant contributions emerge: A) we demonstrate numerically how constrained accessibility can be specified, and doing so brings interpretable units to each accessibility value, units of 'number of potential opportunities for spatial interaction'. B) we provide a clarifying framework for understanding the concept of 'potential' through constraints. Namely, demonstrating accessibility's maximum potential is captured by the the 'total constraint' (constrained version of Hansen-type measure), a restricted amount of potential is captured by the 'single constraint' (i.e., competition, such as the 2SFCA), and potential is fully restricted in the 'double constraint', effectively restricting potential such that this case is equivalent to access or simply spatial interaction. C) Demonstrate how the two most popular accessibility measures, the Hansen-type measure (non-competitive) and the 2SFCA (competitive), link into the  family of accessibility measures.

In fewer words: this work makes an important contribution to accessibility literature, by providing accessibility with a more interpretable, clarified and unified definition through the introduction of constraints from spatial interaction modelling, that reintroduce units to enhance interpretability of raw values, clarify the concept of potential, and loops the most popular forms of accessibility measures into the family. -->

As industrializing cities grew rapidly in the early nineteenth century, traffic congestion became a major issue, prompting the development of transportation planning focused primarily on mobility. In this newly founded practice, access to destinations was treated as a by-product of movement. Following the rise of the automobile and significant investments in transportation infrastructure after World War II, this mobility-oriented approach has led to some problematic outcomes. Specifically, the car became seen as the ultimate mobility tool, helping to foster the development of low-density, single-use residential neighborhoods and entrenching an automobility mono-culture within transportation systems [@miller_collaborative_2011; @lavery_driving_2013]. Decades of planning for this automobility mono-culture, still often characterized by road and highway expansion, have been marked by increased travel costs and environmental burdens, with limited impact on enhancing the ease with which people can reach destinations [@farber_running_2011; @handyACCESSIBILITYVSMOBILITYENHANCING2002; @paez_healthcare_2010]. In response, transportation researchers have increasingly advocated for the adoption of accessibility as a planning criterion, in contrast to traditional mobility-oriented transportation planning approaches which translate into indicators that benchmark movement (e.g., vehicle kilometres traveled, intersection through traffic, etc.) which are not necessarily linked to improved accessibility [@silvaAccessibilityInstrumentsPlanning2017; @paez_developing_2013; @handy2020; @elgeneidyMakingAccessibilityWork2022].

Accessibility can be seen to represent an evolution of mobility; accessibility may be defined as the "potential of opportunities for [spatial] interaction" [@hansen1959], in contrast to mobility which is simply, movement in itself or 'spatial interaction'. Mobility has been the basis of traditional transportation planning approaches [@ortuzar_2011_modelling]. While accessibility brings a more holistic understanding of transportation and land use systems combined [@handyMeasuringAccessibilityExploration1997], by incorporating both the explicit consideration of destination as opportunity and the concept of _potential_.

The growing interest in accessibility has been accompanied by a boom in scholarly research using different methods and focusing on different research contexts; it has grown to include studies of access to employment [e.g., @karstEvaluationAccessibilityImpacts2003; @grengs2010job; @paez_jobs_2013; @merlin2017competition; @tao_investigating_2020], health care [e.g., @luo2003; @paez_healthcare_2010; @wan2012three; @delamater2013spatial; @boisjoly2017informality; @pereira_2021_geographic], green spaces [@reyesAccessibility2014; @rojas_accessibility_2016; @liang_novel_2024], schools [e.g., @williams_disparities_2014; @romanillosAccessibilitySchoolsSpatial2018; @marques_accessibility_2021], social contacts [e.g., @neutens_human_2007; @farberActivitySpacesMeasurement2012; @farber_2013_social], and regional economic analysis [e.g., @vickermanAccessibility1999; @lopezMeasuring2008; @ribeiro_road_2010; @gutierrez_evaluating_2011] among many other domains of application. In other words, accessibility analysis is used today to broadly understand the potential to reach (or spatially interact with) opportunities that are important to people [@ferreiraReenactingMobilityAccessibility2020]. However, despite its growth in popularity in scholarly works, challenges remain with respect to the more widespread adoption of accessibility in planning practice. Several barriers to bridging the scholarly-practice accessibility gap have been identified. For example, the diversity of accessibility definitions has been flagged by @vanweeAccessible2016, @handy2020, and @kapatsila_resolving_2023. Further, difficulties in the interpretability and communicability of outputs has also been noticed by @geursAccessibilityEvaluationLanduse2004, @vanweeAccessible2016, and @ferreiraReenactingMobilityAccessibility2020.

Furthermore, the adoption of accessibility in planning practice is not necessarily made easier when potential adopters have to contend with a plethora of definitions, each seemingly more sophisticated but less intuitive than the last [@kapatsila_resolving_2023]. At a high level, @geursAccessibilityEvaluationLanduse2004 identify four families of accessibility measures: infrastructure-, place-, person-, and utility-based. Of the place-based family, which is the focus of this paper, the menu has grown to include gravity-based accessibility [e.g., @hansen1959; @pirie_measuring_1979], cumulative opportunities [e.g., @wachs_physical_1973; @pirie_measuring_1979; @ye_spatial_2018], modified gravity [e.g., @SchuurmanMeasuring2010], 2-Step Floating Catchment Areas [e.g., @luo2003], Enhanced 2-Step Floating Catchment Areas [e.g., @luoEnhanced2009], 3-Stage Floating Catchment Areas [e.g., @wan2012three], Modified 2-Step Floating Catchment Areas [e.g., @delamater2013spatial], inverse 2-Step Floating Catchment Areas [e.g., @wang_2sfca_2021], and n-steps Floating Catchment Areas [@liang_novel_2024]. How is a practitioner to choose among this myriad options? What differences in accessibility scores should matter, and how should they be communicated? [see @vanweeAccessible2016; p. 14].

More specifically, this work seeks to address the breath of place-based accessibility's breath of definitions and lack of interpretability by first demonstrating that and mobility-oriented models, such as the commonly used gravity-based accessibility [e.g., @hansen1959] and the spatial interaction model [e.g., @wilson1971], respectively, are rooted in the same spatial interaction modeling foundation. In fact, accessibility can be specified using spatial interaction principles. And a _family of measures_, akin to the family of spatial interaction models introduced in @wilson1971, can be defined using balancing factors that constraint values based on known information about the land-use.

In this way, this work aims to introduce constrained accessibility measures, namely the totally-constrained accessibility measure and the singly-constrained accessibility measure, and demonstrate their connection to the Hansen-type accessibility [@hansen1959], the popular competition approach of the 2-Step Floating Catchment Area (2SFCA) [@shenLocationCharacteristicsInnercity1998; @luo2003], and the concept of market potential [@harris_market_1954; @vickermanAccessibilityAttractionPotential1974]. These Wilson-analogous constraints can also be defined in order of increasing restrictiveness to shed light on the role of _potential_ in accessibility and access. We also aim to demonstrate the intuitive clarity in restoring units to resulting accessibility values; using the constrained measures each zonal value retains interpretable units, namely the 'number of opportunities'. This is in contrast to conventional accessibility measures that yield a value that is in units of some 'opportunities-weighted-by-some-representation-of-travel-friction'.

<!-- In this respect, @wuUnifyingAccess2020's contribution provides a unifying framework to think about accessibility measures, the most general of which they describe as $S_i$ in @eq-access-Lu-Levinson. By considering the key features of accessibility, namely travel-cost $f(c_{ij})$ and the distribution of opportunities $g(O_i)$, @wuUnifyingAccess2020 demonstrate that a majority of the concepts found scattered throughout the accessibility literature can be seen as particular cases of a general accessibility formula (@eq-access-Lu-Levinson). One needs only to judiciously change the way travel costs and opportunities are formulated to derive almost any known place-based accessibility measure. In this way, @wuUnifyingAccess2020 have taken a considerable step towards clarifying the differences between various accessibility indicators. -->

<!-- $$ -->
<!-- S_i \propto \sum_j g(O_j)f(c_{ij}) -->
<!-- $$ {#eq-access-Lu-Levinson} -->

<!-- However, there is room to further clarify another relevant aspect: competition, i.e. when finite resources can be concurrently accessed by many people. @wuUnifyingAccess2020's unifying framework does not appear to accommodate many (or any) of the floating catchment area methods, a popular approach in the accessibility literature that accounts for competition. There have also been more recent developments in the literature: for instance, @soukhovIntroducingSpatialAvailability2023 demonstrate that the introduction of a single constraint is sufficient to turn the general accessibility measure (i.e, @eq-access-Lu-Levinson) into a competitive measure of accessibility in accordance with the 2-Step Floating Catchment Area approach. Moreover, the proportional allocation balancing factors in @soukhovIntroducingSpatialAvailability2023 apply a singly-constraint to the accessibility measure, akin to the constraints introduced within the spatial interaction modelling framework outlined in @wilson1971. -->

<!--Motivated by the challenge of incorporating competition into a unified framework for accessibility -->To narratively reiterate: this paper contends that accessibility research must reconnect with its spatial interaction origins. Particularly, we argue that an important aspect of spatial interaction modelling--namely, constraining the results to match empirical observations--was never effectively reincorporated into accessibility analysis. Empirical constraints were embraced by early spatial interaction literature following the work of @wilson1971, but this stream of literature tended to flow separately from research inspired by @hansen1959' accessibility. The application of @wilson1971's empirical constraints supported the development of various spatial interaction models that remain relevant in research and practice today [@ortuzar_2011_modelling]. However, the same cannot be said of the contemporary accessibility literature, where empirical constraints were not explicitly adopted. We argue that the absence of empirical constraints (and their attendant proportionality constants) has contributed to some of accessibility analysis' interpretability issues; for instance, the fuzziness of insights beyond simple proportional statements like 'higher-than' or 'lower-than' [@millerAccessibilityMeasurementApplication2018]. Moreover, without constraints: we lose track of what accessibility measures, that is, the number of opportunities that can be spatially interacted with. Without a clear sense of measurement units, comparability between accessibility measures, across cities, and transport modes, may also become compromised.

Fortunately, accessibility and spatial interaction modelling literature share common headwaters, and the latter has given careful attention to measurement units and their interpretability. It is by looking to the past that we believe accessibility analysis can newly wade into the future. The primary focus of this work, therefore, is to demonstrate how the empirical constraints used in @wilson1971's family of spatial interaction models can be mapped onto accessibility. Both share similar origins in spatial interaction, and popular accessibility measures can be traced directly back to these roots, offering a solution for reintroducing units to the outputs. In the following section, we begin by tracing the development of accessibility from its origins in spatial interaction, from the Newtonian gravitational expression in @ravensteinLawsMigration1889 through to the seminal accessibility work of @hansen1959. We then present evidence for a narrative highlighting the marked divergence between accessibility and spatial interaction modelling research after the work of @wilson1971. Next, we hark back to @wilson1971's spatial interaction models, and use them to derive a family of accessibility measures based on the total and single constraints. We illustrate various members of this family with a simple numerical example. We then conclude by discussing the uses of these measure and their interpretation.

Using the newly defined family of accessibility measures, this work aims to: (1) demonstrate how constrained accessibility can be specified, and by doing so, bring interpretable units to each accessibility value, units of 'number of potential opportunities for spatial interaction'; (2) provide a clarifying framework for understanding the concept of _potential_ through the introduced constraints; and (3) highlight how the two most popular accessibility measures, the Hansen-type measure (non-competitive) and the 2SFCA (competitive), link into the family of accessibility measures.

@@

To be certain, the expression above, a proportionality, is one of the most famous in all of science. In brief, it states that the force of attraction $F$ between a pair of bodies $i$ and $j$ is directly *proportional* to the product of their masses $M_i$ and $M_j$, and inversely *proportional* to the square of the distance between them $D_{ij}$. Direct proportionality means that as the product of the masses increases, so does the force. Likewise, inverse proportionality means that as the distance increases, the force decreases. @eq-phys-grav-prop, however, does not quantify the magnitude of the force. To do so, an empirical constant, a.k.a. the gravitational constant, is required to convert the proportionality into an equality, ensuring that values of the force $F$ in @eq-phys-grav-prop match the observed force of attraction between masses. In other words, @eq-phys-grav-prop needs to be *constrained* using empirical data. Ultimately, the equation for the force is as seen in @eq-phys-grav, where $G$ is an empirically calibrated proportionality constant:

@@

# Early research on human spatial interaction: from Ravenstein (1889) to Stewart (1948)
<!-- 
SUMMARY: a review of spatial interaction literature from the 1880s to the 1940s. A number of researchers theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $Mi$ and $Mj$ and inversely proportional by their separation distance. This concept was captured with different expressions, but all tie back to the same Newtonian gravity analogy, although not all of them included a proportionality constant in their formulation.
-->

Since the 1880s to the 1940, a number of researchers theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $M_i$ and $M_j$ and inversely proportional by their separation distance. This concept was captured with different expressions, but all tie back to the same Newtonian gravity analogy, although not all of them included a proportionality constant in their formulation.

@@

-   $F$ is the *demographic force*
-   $N_1$ and $N_2$ are the numbers of people of in groups 1 and 2
-   $\mu_1$ and $\mu_2$ are so-called *molecular weights*, the attractive weight of groups 1 and 2
-   $M_1 = \mu_1N_1$ and $M_2 = \mu_2N_2$ are the demographic masses at 1 and 2
-   $d_{12}^2$ is the distance between $1$ and $2$
-   And finally $G$, a constant that Stewart "left for future determination" [-@stewartDemographicGravitationEvidence1948, p. 34]

In addition to demographic force, Stewart defined a measure of the "population potential" of group $2$ with respect to group $1$. In other words, the potential number of people from location $2$ that could visit location $1$, as follows:

$$
V_1 = G\frac{M_2}{d_{12}}
$$ {#eq-stewart-population-potential}

For a system with more than two population bodies, Stewart formulated the population potential at $i$ as follows (after arbitrarily assuming that $G=1$):

$$
V_i = \int\frac{D}{r} ds
$$ {#eq-stewart-population-potential-integral}

\noindent where $D$ is the population density over an infinitesimal area $ds$ and $r$ is the distance to $i$. In @eq-stewart-population-potential-integral, $D\cdot ds$ gives an infinitesimal count of the population, say $dm$, and so, after discretizing space, @eq-stewart-population-potential-integral can be rewritten as:

$$
V_i = \sum_j \frac{M_j}{d_{ij}}
$$ {#eq-stewart-population-potential-sum}

@@

<!-- RP: I say we spell it out as in this eq^, what you think? AS: Maybe! -->

@@

# Hansen's gravity-based accessibility to today {#grav-to-today}

@@

However, Hansen's use of Stewart's *population potential* measure included one crucial omission that afflicts the literature to this day. The omission is that that between @stewartDemographicGravitationEvidence1948 and @hansen1959, the proportionality constant $G$ in @eq-stewart-population-potential vanished. This constant was not explicitly addressed in @hansen1959 and accessibility research continues to evolve without it. Since @hansen1959, accessibility analysis has been widely used in numerous disciplines but, to our knowledge, the proportionality constant has remained implicit or forgotten, with no notable developments to explicitly acknowledge or determine it.

The omission of this constant generates a fundamental problem for the measurement unit of accessibility estimates, which has fundamental implications for the interpretation, communication and comparability of accessibility analysis. Those reading @hansen1959 must recall that @stewartDemographicGravitationEvidence1948 had set the proportionality constant $G$ to 1, with a note that "$G$ \[was\] left for future determination: a suitable choice of other units can reduce it to unity" [p. 34]. In practice, the persistent omission of the constant in accessibility analyses means that $G$ continues to be implicitly set to $1$, even when the fundamental relationship in accessibility is proportionality (e.g., $S_{i} \propto \sum_j g(O_j)f(d_{ij})$) and not equality [for instance, see the formula for accessibility at the top of Figure 1 in @wuUnifyingAccess2020]. The direct consequence is that without a proportionality constant, the units of $S_i$ remain unclear: the unit of "potential of opportunity for interaction" is left free to change as $\beta$ is calibrated. For example, if $c_{ij}$ is distance in meters, it will be number of opportunities per $m^{\beta}$ when $f(c_{ij}) = d^{-\beta}$ but number of opportunities per $e^{-\beta\cdot m}$ when $f(c_{ij}) = e^{\beta\cdot m}$. This undermines the comparability of accessibility metrics with different decay functions, and renders their results difficult to understand and communicate. Hansen-style accessibility, therefore, is better thought of as an ordinal measure of potential that can only be interpreted in terms of higher and lower accessibility, but which has not palpable meaning [@millerAccessibilityMeasurementApplication2018].

@@

On the other side of the Atlantic, Alan G. Wilson was developing related, yet parallel work. In the groundbreaking [@wilson1971], Wilson defined a general spatial interaction model as follows. While accessibility was characterised as an associated concept of 'potential', the primary focus was on modelling spatial interaction:

On the other side of the Atlantic, Alan G. Wilson was developing related, yet parallel work. In his groundbreaking study [@wilson1971], accessibility is acknowledged as an associated concept, though the work's primary focuses on defining an introducing a general spatial interaction model as follows:

@@

The model in @eq-phys-gravity-model posits a quantity $T_{ij}$ that represents a value in a matrix of flows of size $n \times m$, that is, between $i = 1,\cdots, n$ origins and $j = 1,\cdots, m$ destinations. The quantities $W_i^{(1)}$ and $W_j^{(2)}$ are proxies for the masses at $i=1,\cdots,n$ origins and $j=1,\cdots,m$ destinations. Finally, $f(c_{ij})$ is some function of travel cost $c_{ij}$ which reflects travel impedance. In this way, $T_{ij}$ explicitly measures *interaction* in the unit of trips, and the role of $k$ is to ensure that the system-wide sum of $T_{ij}$ represents the total flows in the data. In other words, $k$ is a scale parameter that makes the overall amount of flows identical to the magnitude of the phenomenon being modeled. In other words, it balances the units, in a conceptually similar sense as the gravitational constant in Newton’s Law of Universal Gravitation.

@@

Additional information can be introduced. For example, when information is available about the total number of trips produced by each origin, $W_i^{(1)}$ is represented as $O_i$ and the following constraint can be used:
